Observation of Red Wine Data Set by Renel Garces
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

#loading libraries 

library(ggplot2)
library(dplyr)
library(reshape)
library(MASS)
library(car)
library(GGally)
library(memisc)
library(reshape2)
library(scales)
library(gridExtra)
library(lattice)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

wine <- read.csv('wineQualityReds.csv')

```

  This report includes the examination of the red wine data set.  The main focus
of this report is to find common characteristics that are shared between the 
same quality of wines.  There will also be analysis conducted to observe the
correlation between different variables to wine quality.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

dim(wine)
str(wine)
summary(wine)

```

The data set includes 1599 observations with thirteen different variables.
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = pH), data = wine) +
  geom_histogram(binwidth = .01) +
  ggtitle('Count of pH Observations')

summary(wine$pH)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = citric.acid), data = wine) + 
  geom_histogram(binwidth = .01) +
  xlab('citric acid (g / dm^3)') +
  ggtitle('Count of Citric Acid Observations')

summary(wine$citric.acid)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

#xlim has been placed to ignore the top 5% of the data

ggplot(aes(x = residual.sugar), data = wine) + 
  geom_histogram(binwidth = .1) +
  xlim(c(0, quantile(wine$residual.sugar, .95))) +
  xlab('residual sugar (g / dm^3)') +
  ggtitle('Count of Residual Sugar Observations')

summary(wine$residual.sugar)

```

A limit on the value of residual.sugar was placed in order to focus on the bulk
of the data.  Most wines have a sugar level around 2 although there are some
outliers having a sugar level up to 15, causing the data to be skewed to the 
right.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = chlorides), data = wine) + 
  geom_histogram(binwidth = .001) + 
  xlim(c(0, quantile(wine$chlorides, .95))) +
  xlab('chlorides g / dm^3') +
  ggtitle('Count of Chloride Observations')

summary(wine$chlorides)

```


A limit was placed on the value of chlorides to focus on the bulk of the data 
and ignore the outliers.  The top 5% of values are ignored in this plot.  The
observations of chloride seem to be the most normally distributed variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = density), data = wine) + 
  geom_histogram(binwidth = .0001) +
  xlab('density (g / cm^3)') +
  ggtitle('Count of Density Observations')

summary(wine$density)

```

The density variable appears to be normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = alcohol), data = wine) + 
  geom_histogram(binwidth = .1) +
  xlab('alcohol (% by volume)') +
  ggtitle('Count of Alcohol Observations')

summary(wine$alcohol)

```

The variable alcohol is skewed towards the right, with most observations having an alcohol rating of 9.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = sulphates), data = wine) + 
  geom_histogram(binwidth = .01) +
  xlim(c(0, quantile(wine$sulphates, .99))) +
  xlab('sulphates g / dm3') +
  ggtitle('Count of Sulphate Observations')

summary(wine$sulphates)

```

The sulphate variable was the most heavily skewed variable.  The top one percent
of data is ignored in this plot to avoid seeing overly dispersed data.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = quality), data = wine) + 
  geom_histogram(binwidth = .1) +
  ggtitle('Count of Quality Observations')

summary(wine$quality)

```


````{r echo=FALSE, message=FALSE, warning=FALSE}
# group all wines into high or low quality

condition <- (wine$quality > 5)
wine$qual_cat <- ifelse(condition, 'high', 'low')

highs <- subset(wine, qual_cat == 'high')
lows <- subset(wine, qual_cat == 'low')

qplot(data=wine, x=qual_cat, xlab='quality category',
      ylab='count', main='Wine by Quality Category')

summary(wine$qual_cat)

```

# Univariate Analysis

### What is the structure of your dataset?

The structure of the data set includes 1599 observations with 13 different
variables.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in this report is to view the trends among 
different wines grouped by quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I believe that there will be a negative relationship between acidity(pH) and 
quality as it's generally thought that higher quality wines have lower the pH.
I also hope to find that the wines containing the highest and lowest residual
sugar values will be classified under lower quality ratings as I presume that
those wines will be classified as low quality due to the imbalance of sweetness.

### Did you create any new variables from existing variables in the dataset?

No

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The only normally distributed variables were density and pH.  All other 
variables appeared to be long-tailed and skewed towards the right.

In order to tidy the data, I adjusted the binwidths on plots with variables such
as pH, citric.acid, and residual.sugar because the data points consisted of data
that was calculated to the tenth or hundredth decimal place.  This was to 
prevent rounding of the data that would cause the plots over cluster the data
and make it difficult to see the variations in the distributions. I also limited
the values along the x-axis on some of the plots in order to focus the plots 
towards the bulk of the data and ignore the outliers.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

#Creating a new data frame that will contain mean values for each quality rating

wine_qualities <- summarise(group_by(wine, quality), 
          mean_density = mean(density),
          median_density = median(density),
          mean_pH = mean(pH),
          mean_sulphates = mean(sulphates),
          mean_sugar = mean(residual.sugar),
          mean_chlorides = mean(chlorides),
          mean_citric = mean(citric.acid),
          n = n())
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

sevens <- subset(wine, quality == 7)
eights <- subset(wine, quality == 8)

ggplot(wine, aes())  +
  geom_histogram(data = sevens, aes(x = citric.acid, colour=factor(quality)),
                 alpha = .2, binwidth = .01) +
  geom_histogram(data = eights, aes(x = citric.acid, colour=factor(quality)),
                 alpha = .2, binwidth = .01) +
  scale_colour_manual(values = c("red", "blue")) +
  xlab('citric acid (g / dm^3)') +
  ggtitle('Citric Acid Count for 7 and 8 Quality Wines')
  
```

This plot shows that the highest quality wine (8) show no particular trends of 
citric acid levels, while most of the second highest quality of wines (7) have 
an citric acid level of 0 or between .3 and .5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wine, aes(x = citric.acid, y = pH)) +
  geom_point(alpha = 1/5) +
  xlab('citric acid (g / dm^3)') +
  ggtitle('Citric Acid vs pH')

cor.test(wine_qualities$mean_citric, wine_qualities$quality)
```

There is a postive relationship between a wine's citric acid level and overall 
acidity level (or negative in terms of pH level).  

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_qualities, aes(y = mean_sugar, x = quality)) +
  geom_bar(stat='identity', position='dodge') +
  ylab('mean_sugar (g / dm^3)') +
  ggtitle('Mean Sugar vs Quality')

cor.test(wine_qualities$mean_sugar, wine_qualities$quality)
```

There's very low correlation between a wine's residual sugar levels and its quality rating.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wine_qualities, aes(x = mean_chlorides, y = quality)) +
  geom_line() +
  xlab('mean chlorides g / dm^3') +
  ggtitle('Mean Chlorides vs Quality')

cor.test(wine_qualities$mean_chlorides, wine_qualities$quality)
```

This plot shows that as chloride content decreases, quality rating increases.  
There is a noticeable dip for moderate quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_qualities, aes(x = mean_density, y = quality)) +
  geom_line() +
  xlab('mean_density (g / cm^3)') +
  ggtitle('Mean Density vs Quality')

cor.test(wine_qualities$mean_density, wine_qualities$quality)

```

This plot shows that higher quality wines are more likely to have a lower 
density.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_qualities, aes(x = mean_pH, y = quality)) +
  geom_line() +
  ggtitle('Mean pH vs Quality')

cor.test(wine$pH, wine$quality)

```

This plot shows that as quality decreases, pH also increases but very slightly.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_qualities, aes(x = mean_sulphates, y = quality)) +
  geom_line() +
  xlab('mean_sulphates g / dm3') +
  ggtitle('Mean Sulphates vs Quality')

cor.test(wine_qualities$mean_sulphates, wine_qualities$quality)

```

This plot shows the positive relationship between quality and mean_sulphates.
This plot was also one of the only ones to not show the moderate quality wines
breaking the trend between the high vs low quality wines.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

After analyzing different relationships between different variables in the data
set, it has been discovered that a lot of the variables were weakly correlated
to the variable of quality including; acidity, sugar, chloride, density,
sulphates unless you correlate the mean value of each variable in relation
to quality.  Mean residual sugar showdd the weakest correlation to quality while
mean sulphates showed the strongest correlation to quality.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed the correlation between citric acid content and overall pH.  It was
the highest positive correlation observed at 0.97.  It was also interesting to
see how little sugar had to do with the quality ratings.


### What was the strongest relationship you found?

The strongest relationship found was the correlation between sulphates and
quality rating at 0.9859.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}

ggplot(data = wine_qualities, aes(x = mean_pH, y = mean_sugar)) +
  ylab('mean_sugar (g / dm^3)') +
  geom_point(aes(color = factor(quality))) +
  ggtitle('Sugar and pH Composition of Wine Qualities')

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# add more statistical sulphate variables

wine_qualities <- summarise(group_by(wine, quality), 
          mean_density = mean(density),
          median_density = median(density),
          mean_pH = mean(pH),
          mean_sulphates = mean(sulphates),
          mean_sugar = mean(residual.sugar),
          mean_chlorides = mean(chlorides),
          mean_citric = mean(citric.acid),
          max_sulphate = max(sulphates),
          min_sulphate = min(sulphates),
          median_sulphate = median(sulphates),
          n = n())

ggplot(data = wine_qualities, aes()) +
  geom_line(aes(y = mean_sulphates, x = quality, color = 'mean')) +
  geom_line(aes(y = median_sulphate, x = quality, color = 'median')) +
  ylab('sulphates g / dm3') +
  ggtitle('Mean and Median Sulphates vs Quality')

```

This plot's purpose is to illustrate the mean composition of different quality
wines in terms of acidity and sugar.  There appears to be no trend between pH
and sugar composition among the different quality of wines.


```{r echo=FALSE, message=FALSE, warning=FALSE}

#Selecting fewer of the wine qualities to not overcrowd the plot

sub <- subset(wine, quality %in% c(3, 5, 7))

ggplot(data = sub, aes(x = alcohol, y = residual.sugar), binwidth = .001) +
  geom_point(aes(color = factor(quality))) +
  xlab('alcohol (% by volume)') +
  ylab('residual sugar (g / dm^3)') +
  ylim(c(1, quantile(sub$residual.sugar, .95))) +
  ggtitle('Sugar and Alcohol Balance of Low/Med/High Grade Wines (3, 5, 7)')

```

This plot shows that sugar does not increase as alcohol content increases.  It
also shows that higher quality wines tend to have more alcohol than lower 
quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = wine_qualities, aes(x = mean_chlorides, y = mean_pH)) +
  geom_line(aes(color = quality)) +
  xlab('mean_chlorides g / dm^3') +
  ggtitle('Chloride and pH Composition of Wine Qualities')

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The relationships observed in the multivariant analysis included the observation
of sugar and pH content among different qualities of wine, the combination of 
alcohol and sugar content among the different wine qualities, and the mean
chloride and pH level among different wine qualities.  The main takeway
from the previous three plots is that it is difficult to see a distinction
between sugar content and the quality of wine even when paired with other 
variables such as pH, chloride content, and alcohol content.

### Were there any interesting or surprising interactions between features?

When all three plots are examined together it is revealed the residual sugar
content would play little if any role being able to predict a wines quality.  It
was also interesting to see the high quality wines and low quality wines both
shared the characteristic of containing relatively high amounts of sugar.  The
differentiating factor may be the sugar content in higher quality wines is 
balanced out by the higher alcohol content.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

N/A

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE}

# group all wines into high or low quality

condition <- (wine$quality > 5)
wine$qual_cat <- ifelse(condition, 'high', 'low')

highs <- subset(wine, qual_cat == 'high')
lows <- subset(wine, qual_cat == 'low')

ggplot(wine, aes())  +
  geom_histogram(data = highs, aes(x = citric.acid, colour=factor(qual_cat)),alpha = .05, binwidth = .01) +
  geom_histogram(data = lows, aes(x = citric.acid, colour=factor(qual_cat)), alpha = .05, binwidth = .01) +
  xlim(quantile(lows$citric.acid, .05), quantile(lows$citric.acid, .95)) +
  xlim(quantile(highs$citric.acid, .05), quantile(highs$citric.acid, .95)) +
  ylim(0, 35) +
  scale_colour_manual(values = c("red", "blue")) +
  xlab('citric acid (g / dm^3)') +
  ggtitle('Citric Acid Count for High and Low Quality Wines')

```

### Description One

This plot shows the distribution of different citric acid levels between low
quality wines and high quality wines.  The low quality wines are wines rated 
from 3-5 while the high quality wines are rated 6-8.  The distribution of 
low quality wines appear to be slightly more weighted to the lower citric acid
levels while the higher quality wines appear to be slightly more weighted to the
higher citric acid levels.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

#Selecting fewer of the wine qualities to not overcrowd the plot

sub <- subset(wine, quality %in% c(3, 5, 7))

ggplot(data = sub, aes(x = alcohol, y = residual.sugar), binwidth = .001) +
  geom_point(aes(color = factor(quality))) +
  xlab('alcohol (% by volume)') +
  ylab('residual sugar (g / dm^3)') +
  ylim(c(1, quantile(sub$residual.sugar, .95))) +
  ggtitle('Sugar and Alcohol Balance of Low/Med/High Grade Wines (3, 5, 7)')

```

### Description Two

This plot was created to see if sugar and alcohol balance had anything to do 
with a wine's quality.  This plot reveals sugar is irrelevent to a wines quality
given its alcohol content.  The plot also reveals that higher quality wines tend
to contain more alcohol relative to lower quality wines.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

# add more statistical sulphate variables

wine_qualities <- summarise(group_by(wine, quality), 
          mean_density = mean(density),
          median_density = median(density),
          mean_pH = mean(pH),
          mean_sulphates = mean(sulphates),
          mean_sugar = mean(residual.sugar),
          mean_chlorides = mean(chlorides),
          mean_citric = mean(citric.acid),
          max_sulphate = max(sulphates),
          min_sulphate = min(sulphates),
          median_sulphate = median(sulphates),
          n = n())

ggplot(data = wine_qualities, aes()) +
  geom_line(aes(y = mean_sulphates, x = quality, color = 'mean')) +
  geom_line(aes(y = median_sulphate, x = quality, color = 'median')) +
  ylab('sulphates g / dm3') +
  ggtitle('Mean and Median Sulphates vs Quality')

```

### Description Three

This plot illustrates the strongest correlation in the data set which is the
relationship between wine quality and sulphate amount.  I added the median line
in the graph to provide more statistical evidence that sulphate is strongly,
positively correlated with wine quality.

------

# Reflection

Overall this project opened to my eyes to how trends may not be as easily
visible by looking at the raw data.  Trends may be hidden because of the skewing
caused by the over dispersion of data, the overcrowding of data points, or 
approaching the data itself in the wrong way.  At first I struggled to find any
meaningful relationships when I looked at the data with minimal manipulation. 
Once I've gotten to seperating the data set into different groups by wine 
quality and looking at the means of the other variables, it became easier to see
trends and relationships.  It was suprising to see that, despite my prediction,
there was no trend in combinations of variables of pH and sugar, acidity and
sugar, and alcohol and sugar when it came to predicting wine quality.  I thought
that higher wine qualities would share similar characteristics of a balance of 
sweet/salty, sweet/acidic, or sweet/alcohol.  I was wrong.  In the future, 
one may further examine this data set by taking an even sample from each wine
quality group and conduct analysis of the variables to make up for the unequal
representation of each wine quality in the data set.
